Supplementary material

### 2021-12-21 analysis/paper/paper.Rmd
Minor edits

The shore-bound settlement is believed to follow both from a resource perspective and from one of movement and communication, which could be more efficient on waterways. The is thus also extended to the hinterland and the inland, where sites are to be located along rivers and by inland lakes. This picture is to take a dramatic turn around 2400 BCE, where the introduction of Neolithic package associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline. That is not to say that aquatic resources were no longer exploited, but settlements and tool-production areas were no longer

Based on the generally accepted premise that most Stone Age sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which are often guided by a both a digital and mental reconstruction of past sea-levels. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the possible time-span provided by the typological indicators are too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea-level was higher, it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are fewer, and are typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore-bound site location is largely based on a mosaic of smaller scale investigations.

@@

Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @solheim2020; @åstveit2018]. This approach has a couple of limitations. First of all, the displacement curves are commonly applied directly to larger study areas, with only a couple of studies having taken the variable uplift-rates into account when performing this comparison [e.g. @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of the relation between site and shore line, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating. Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.

@@

All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a digital research compendium following @marwick2018. The dataset is slightly more extensive than what is directly utilised in the study, and includes excavated sites without ^14^C-dates, as well as ^14^C-dates to periods other than the Stone Age.

To get at the relationship between sites and contemporaneous shoreline, the analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

The employed shoreline displacement curves are all based on the so-called isolation basin method [@romundset2011]. This involves extracting cores from a series of basins located beneath the marine limit situated on bedrock in a constrained area of the landscape, and dating the transition from marine to lacustrine sediments in the cores. Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point. The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets, the adjustment to a common isobase based on the direction and gradient in rebound intensity, to name but a few [e.g. @romundset2011; @romundset2019]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications (see above).

The archaeological data compiled for the analysis consists of all excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)). These number a total `r nrow(sites)` sites. Of these `r length(unique(rcarbon$site_name))` sites are associated with the total of `r nrow(rcarbon)` radiocarbon dates. In turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability. These sites form the basis for the analysis. Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History---the institution responsible for archaeological excavations in the region. In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.

@@

The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2019]. After evaluating the options I decided to use 10m resolution DTM rather than the higher-resolution 1m version. In addition to resulting is less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by. The 10m DTM is a down-sampled version of the 1m resolution DTM, which is based on aerial laser scanning using a minimum of two and up to five points per m^2^ [@mapping2019].

@@

To take account of the gradient in the isostatic rebound, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done by for each site first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the distance between the isobases. To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. These values were then subtracted from the dsiplacement curve located to the south-west of the site. The result of this process is shown for an example site in Figure \@ref(fig:example-1).

For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase. In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal. Multiple phases at a single site were treated as independent of each other.

@@

Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulations were performed. A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)). This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horizontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The topographic distance between site and sea was also found by measuring the shortest distance while taking into account the slope of the terrain on the DTM. This was done using the topoDistance package for R [@wang2019]. The topographic distance was measured between site polygon and the horizontally closest point on the shoreline. The shortest topographic path was found using the Queen's case neighbourhood of eight cells [@conolly2006]. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygon completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values where therefore retained.

@@

The result of running the simulations across all sites are given in Figure \@ref(fig:results). Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, most of the sites have been situated close to the shoreline when they were in use. When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline, as could be expected. To what degree these do in fact represent contamination is discussed further below. However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. It is interesting to note the distance for some of the earliest sites, which appears somewhat high. This, however, can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, they appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance. This corresponds well with the literature, where the introduction of the Neolithic proper is seen as occurring around this time [@prescott2020; @solheim2021]. With the transition from a predominantly hunter-gatherer lifeway to early agro-pastrolism, settlements in coastal areas are to be increasingly withdrawn from the shoreline. Given this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE.

@@

Some sources of likely variation and uncertainty have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error... Thirdly, the tidal range has not been considered...  Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled. And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site

@@

Here the approach was used to inform the question of the distance between sites and shorelines, but this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore line, and the uncertainty inherent in this reconstruction.

>>>>

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from a resource perspective and from one of movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes. This general picture is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, where the introduction of the Neolithic proper inolved a. associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@solheim2018; @prescott2012]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods.

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which are often guided by a both a digital and mental reconstruction of past sea-levels. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if it would have been located on an island when the sea-level was higher, it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are fewer, and are typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore-bound site location is largely based on a mosaic of smaller scale investigations.

@@

Other previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @solheim2020; @åstveit2018]. This approach has a couple of limitations. First of all, the displacement curves are commonly applied directly to larger study areas, with only a couple of studies having taken the variable uplift-rates into account when performing this comparison [e.g. @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating. Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.

@@

All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a digital research compendium following @marwick2018. The dataset is slightly more extensive than what is directly utilised in the study, and includes excavated sites without ^14^C-dates within the study area, as well as ^14^C-dates to periods other than the Stone Age.

To get at the relationship between sites and the contemporaneous shoreline, the analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

The employed shoreline displacement curves are all based on the so-called isolation basin method [@romundset2011]. This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores. Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point. Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located a significant distance from a common isobase perpendicular to this gradient. The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets, the adjustment to a common isobase, to name but a few [e.g. @romundset2011; @romundset2019]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications (see above).

The archaeological data compiled for the analysis consists of all excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)). These number a total `r nrow(sites)` sites. Of these, `r length(unique(rcarbon$site_name))` sites are associated with the total of `r nrow(rcarbon)` radiocarbon dates. In turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability. These sites and ^14^C-dates form the basis for the analysis. Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History---the institution responsible for archaeological excavations in the region. In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as that defined during initial survey, area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.

@@

The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2019]. After evaluating the options I opted for the 10m resolution DTM rather than the higher-resolution 1m version. In addition to resulting in considerably less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by. The 10m DTM is a down-sampled version of the 1m resolution DTM, which is based on aerial laser scanning using a minimum of two and up to five points per m^2^ [@mapping2019].

@@

To take account of the gradient in the isostatic rebound, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done for each site by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the distance between the isobases. To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. These values were then subtracted from the dsiplacement curve located to the south-west of the site. The result of this process is shown for an example site in Figure \@ref(fig:example-1).

For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase. In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

@@

Armed with a likely date range for the occupation(s) of each site, an estimated trajectory of relative sea-level change at that location, and a DTM edited to remove substantial modern disturbances, the simulations were performed. A single simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase of a site (Figure \@ref(fig:example-2)). This year then has a corresponding likely elevation range for the contemporaneous shoreline from which an elevation value was drawn uniformly, using intervals of 5cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horizontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The topographic distance between site and sea was also found by measuring the shortest distance while taking into account the slope of the terrain on the DTM. This was done using the topoDistance package for R [@wang2019]. The topographic distance was measured between site polygon and the horizontally closest point on the shoreline. This means that the distance is not necessarily measured as the closest topographic distance to the shoreline, but rather as the shortest topographic path to the horizontally closest point on the shoreline. This significantly reduced the computational cost of the analysis, and is deemed unlikely to have a considerable impact, given the distances considered. The shortest topographic path was found using the Queen's case neighbourhood of eight cells [@conolly2006]. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygon completely contain the site, the horizontal and topographic distance measures were made negative, and the vertical distance was instead measured to the highest point on the site polygon. While it is safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure, and might also help identify discrepancies in displacement data or radiocarbon dates. Negative values where therefore retained, with the exception of for the sites Gunnarsrød 5, Pjonkerød R1 and , where the negative values are believed to result from modern disturbances in the DTM rather than the ^14^C-dates or displacement curves (see supplementary material for more details).

@@

The result of running the simulations across all sites are given in Figure \@ref(fig:results). Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, most of the sites have been situated close to the shoreline when they were in use. When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline, as could be expected. To what degree these do in fact represent contamination is discussed further below. However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. It is interesting to note the distance for some of the earliest sites, which appears somewhat high. This, however, can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, they appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance. This corresponds well with the literature, where the introduction of the Neolithic proper is seen as occurring around this time [@prescott2020; @solheim2021]. With the transition from a predominantly hunter-gatherer lifeway to early agro-pastrolism, settlements in coastal areas are to be increasingly withdrawn from the shoreline.

The negative values around originate from the sites Løvås 1 and 2. These are well-dated sites situated in a relatively undistributed area of the landscape. Thus, while there is a danger of circularity of having archaeological sites inform a reconstruction of the shoreline displacement, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper bound, as they would not have been in use when located beneath sea-level. Furthermore, as the Løvås sites are located at elevations from where the local displacement curve is interpolated between isololation basins at different elevation, it would seem that the sites represent a case where the archaeological material indicates discrepancies in the geological reconstructions of shoreline displacement in the area.

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. Following the general and aggregative result of this, one recommendation could be that given no other information, the method of shoreline dating should be applied with an uncertainty of 0-15m offset to the Stone Age site elevation relative to the displacement curves, following the median vertical distance and associated interquartile range. This would naturally be best applied when combined with a more comprehensive consideration of the site location, which could lead one to reduce this uncertainty, given for example surroundings where it would be reasonable to assume that a certain sea-levels would be unlikely in light of the surroundings of the site.

@@

Some sources of likely variation and uncertainty have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error... Thirdly, the tidal range has not been considered...  Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled. And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site. These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method. These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.

@@

The most immediate contribution of this paper is a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline, which has largely been verified. Furthermore, the typical application of shoreline dating as an approximate method where date ranges are often for example rounded off to 200-year intervals, means that most previous research based on the method is given support by the findings reported here. However, the findings also suggest that simply using the method as a terminus ante quem for the occupation of the sites is, at least in my view, overly conservative and would stand to unnecessarily loose important chronological information. Attempting to move these considerations further through an explicit quantification has here resulted in a concrete recommendation of applying an offset of 0-15m when shoreline dating sites where no other information is available.

This quantification was achieved by means of Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty. Here the approach was simply used to inform the question of the distance between sites and shorelines. However, this method can be extended to a wide range of use-cases where one needs to visualise and both quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

### 2021-12-21 analysis/supplementary-materials/supplementary.Rmd
Minor edits

Nordby 52 is a rock-shelter tat has been visited throughout prehistory [@gjerpe2008]. Clear typological date to both of the Stone Age periods defined by the ^14^C-dates treated here. The site is located by where the highway runs today (see also Nordby 1 above). Edit appears to have been successful.

@@

Pjonkerød R1. Situated by a quarry today which has been fairly successfully edited. However, as the few negative values across the results are likely to result from the interpolation process and the fact that this is a quite large area, and not issues with the displacement curves or ^14^C-dates, the negative values were forced to zero for the aggregative analsysis.

@@

>>>>

Nordby 52 is a rock-shelter that has been visited throughout prehistory [@gjerpe2008]. Clear typological date to both of the Stone Age periods defined by the ^14^C-dates treated here. The site is located by where the highway runs today (see also Nordby 1 above). Edit appears to have been successful.

@@

The lithic inventory from Pjonkerød R1 is consistent with the ^14^C-date [@carrasco2015]. The site is situated by a gravel pit today, which has been fairly successfully edited. However, as the few negative values in the results are likely to result from the interpolation process and the fact that this is a quite large area, as opposed to issues with the displacement curves or ^14^C-dates, the negative values were forced to zero for the aggregative analysis.

@@

Prestemoen 1 has a lithic inventory consistent with the ^14^C-dates [@persson2014]. The site is situated by a gravel pit that is clearly visible on the DTM to the north-east of the site. This was not edited as it has not impacted the simulated distance between site and sea. The landscape surrounding the site would, however, have been substantially different in the Mesolithic [@persson2014, fig. 10.20].
